On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2023-09-18 Manuscript/manuscript.tex
finished going through remainder of text; removed resolved

\author{Owen Fiore\\
%   \href{mailto:owen.fiore@uconn.edu}
% {\nolinkurl{owen.fiore@uconn.edu}}\\
  Elizabeth Schifano\\
  Jun Yan\\[1ex]
  Department of Statistics, University of Connecticut\\
}
\date{}

@@

Devon Allen, an alumnus of the University of Oregon in Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field Championships also in Eugene 
in late June, Allen was poised to compete in the 2022 World Track and 
Field Championships in August, which would also be held in Eugene, Oregon.
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
His disqualification in the final heat by reacting in 0.099 seconds, just 0.001
seconds faster than the 0.1 threshold set by the International Association of
Athletics Federations (IAAF), was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification
(\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}).
The crowd erupted with jeers
when they realized that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete. %and been 
%spared waiting until the next World Track and Field Championships.
%\eds{He is not competing this year, correct? If yes, we may need to rephrase.}
%\of{Correct}
%\eds{removed because he did not compete this year, so did not actually have
% to wait, after all.} 
\eds{ If it is true he ``retired'' as a runner to play football, we could 
instead say something like
\textit{Had Allen reacted just 0.001 seconds slower he would have been allowed 
to compete before retiring from running to play football.}}
\of{I don't think he retired, he was competing in races as recently as this past
June.}

@@

The rest of this paper is organized as follows. Section~\ref{sec:Data} details 
how the two types of data were collected, as the signed-rank test as described
in Section~\ref{sec:rank} uses slightly different data than the generalized 
linear mixed-effect model (GLMM) that is further developed in 
Section~\ref{sec:glmm} Section~\ref{sec:Results} provides the results of the two
analyses and ends by questioning the reaction time barrier.  Lastly, 
Section~\ref{sec:concludingremarks} discusses the impact and limitations of the
paper.

@@

There are two types of data that are explored throughout this paper.  For the
signed-rank tests described in Section~\ref{sec:rank}, the data comes from 
athletes who competed in the 2022 World Championships and another competition 
(2019 World Championships, 2023 World Championships, or 2022 national 
championships). For the GLMM described in Section~\ref{sec:glmm}, data comes
from every world championships from 1999 to 2023. This investigation started 
shortly after the 2022 World Championships, but as the 2023 data has now become 
available, we are pleased to include it in both of our analyses, and it did not
significantly change our results.

@@

\subsubsection{2022 National Competitions}\label{sec:datanational}
From June 23-26, 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World 
Championships being held at the same venue in August. Thus, we can establish a 
baseline for the four United States athletes who competed in at least one round 
of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen.   
Every athlete reacted faster in all the World Track and Field Championships 
races compared to the United States Track and Field Championships races. The 
dataset of the four United States athletes is too small to perform any 
meaningful statistical analysis, but the data was expanded to include reaction 
times for 11 more athletes who competed at other national competitions (many of which were 
held from May-July of 2022). This data was difficult to find, as the data is not
centrally located but instead found on different national websites (often 
recorded in the national language without translation). 
\of{I don't know how to say this}. \eds{is my change correct?} \of{No I was able
to use a lot even though I couldn't actually read the page. I just tried to find
110 meter hurdle and then look at a column for reaction times which are always from
roughly 0.1 to 0.18 seconds.} \eds{is it correct now?} \of{Yes}

@@

%\eds{Can you show some descriptive statistics for this data?  Total number of 
%clusters? Average number of international vs national competitions across 
%athletes/clusters?}
%\eds{The following paragraph still needs to be smoothed (the paragraph that was 
%here and the paragraph that was in the intro still need to be integrated into 
%one (or two) cohesive paragraph(s) that accurately describe the two types of 
%data that will be used for two rank-based analyses.}
%\of{Does this work?}
%\eds{Yes, but you still need figure or descriptive statistics for this data.}

@@

While it is theoretically possible that athletes got faster 
between 2019 and 2022 or 2022 and 2023, it is important to 
remember that these are some of the fastest sprinters on the planet, and thus
improving on elite times is not easy.  Once again we repeat a similar procedure
as in Section~\ref{sec:datanational}; we assign each athlete to be a cluster and
within each cluster label times to be from either the 2019, 2022, or 2023 World
Championships. When comparing 2019 and 2022 we look at data from 15
athletes, with two clusters of size two, three clusters of size three, five 
clusters of size four, two clusters of size five, and three clusters of size six.
When comparing 2022 and 2023 we look at data from 22 athletes, with seven 
clusters of size two, three clusters of size three, six clusters of size four,
and six clusters of size five.  
%Although 2022 is presented three times in
%Figure~\ref{fig:RankScatterplots}, the subset of athletes differs each time.
%There may be athletes who competed in 2019 and 2022 and had their national
%championships data available, but they did not compete in 2023, and thus their
%times appear only in the top two graphs. \eds{top and bottom graphs?}
\eds{also, you will likely need to make Fig 1 smaller as the caption runs off
the page (even before my edits).}
\of{It should be smaller now}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics{RankScatterPlots}
  \caption{The top graph shows the reaction times of athletes who competed at 
	least once each at the 2019 and 2022 World Championships. The middle graph 
	shows the reaction times of athletes who competed at least once each at the 
	2022 and 2023 World Championships. The bottom graph shows the reaction times 
	of athletes who competed at least once at their country's
  2022 national championships and the 2022 World Championships.
	\eds{check x-axis labels in the bottom plot: last three should be 2022?}
  \of{I don't think so. Its a comparison of 2022 national vs international so
  I abbreviated it as Nat. and Inter. is that not ok?}}
  \label{fig:RankScatterplots}
\end{figure}

Figure~\ref{fig:RankScatterplots} shows the data that will be analyzed
with the rank-based methods (described in Section~\ref{sec:rank} with results in 
Section~\ref{subsec:Results_Rank}). In each of the three graphs, the reaction 
times for athletes who competed
in the 2022 finals \eds{World Championships?} are compared to their reaction
times at another at which they competed.  
It is worth noting that each instance of the 2022 data is different, because
those who competed at both the 2019 and 2022 World Championships will be
different athletes than those who competed at both the 2022 and 2023 World
Championships. This data was only for athletes who competed
in at least one race at each championship, however it is possible for athletes
to have competed in one race in 2019 and two races in 2022. 
\eds{did we not already cover the last two sentences in the previous paragraph, 
or are you trying to say something different here?} \of{yes you are correct. I
commented out the other one because I think this one flows better.}  The 2022 World
Championship Heats and Semifinals both had on average lower reaction times than
their respective 2019 races, and the 2022 finals were substantially faster. It
is also worth noting is that the fastest reaction time in the 2022 World
Championships Finals and Semifinals both came from Devon Allen, but the 
difference between 0.101 and 0.099
seconds was the difference in what caused him to be disqualified. 
%\eds{Please reorganize the labels so they are either in one of the following 
%orders:
%2019 Heat, 2019 Semi, 2019 final, 2022 Heat, 2022 Semi, 2022 final  OR
%2019 Heat, 2022 Heat, 2019 Semi, 2022 Semi, 2019 final, 2022 final.  
%(I am not sure which is better, but perhaps the latter?)}
%\of{I re-worked this graph completely, please let me know what you think.  For
%each graph I currently list them in chronological order: so 2022 then 2023 but
%I list 2022 second when comparing it to 2019/rank comparison.  Please let me
%know your thoughts on this and if I should list 2022 and then 2023 for visual
%consistency.}
%\eds{I think the graphs look good}

@@

Data was taken from the World Athletics 
and covers the men's 110 meter hurdles from 1999 to 2023.
Although women's data is available, we chose not to include women's data in the
analyses to follow due to the numerous papers citing differences in men and
women's reaction times \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender}.
\eds{you should mention the data source earlier, possible in Section 2 
before Section 2.1 if it applies to Section 2.1.1, and also the sources of the 
national data for 2.1.2 (i.e., pointing to the info in Appendix).}
\of{I want to emphasize the differences in the data so I wanted to try and
compartmentalize everything that was different. We chose not to include women's
data for the GLMM because of cited papers and chose not to include women's
data for the signed-rank because the results were not highly significant. So if
I want to talk about that in section 2, I feel like I would have to explicitly
justify the exclusion of the women's data which then goes back to the
conversation above. Also Prof Yan recommended I remove the appendix, so now the
sources are in the Data README.}
For the creation of the GLMM model we focus on the reaction times during the 
semifinal and final heats only, as reaction times from preliminary heats are 
often not as fast as those in later heats \citep[e.g.,][]{zhang2021correlation}.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics{Boxplot}
  \caption{The reaction times from 1999 to 2023 for the men's 110 meter hurdle.}
	%\eds{there is only one plot here, not two. If you are keeping only one, the
	%text and caption need to be modified accordingly}}
  %\of{Resolved}
  \label{fig:Boxplot}
\end{figure}

@@

With our small sample size, the asymptotic normal distribution may not be
reliable, so we use 1000 random permutations to simulate the null distribution
of the test statistic.  
\eds{simulate the null distribution of the test statistic?}
\of{Resolved}
%\eds{I do not understand what is meant in the text after the colon in the 
%previous sentence.  Please rephrase.}
%\of{I'm not sure what that was, I removed it now.}
This method is available from the \texttt{clusWilcox.test()} function
with \texttt{method = `ds'} (for \underline{D}atta and \underline{S}atten) and 
\texttt{exact = TRUE} from R package
\texttt{clusrank} \citep{jiang2020wilcoxon}.

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the we chose a gamma mixed-effect model as our 
final model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one capturing the venue effect in
each year and the other capturing the heat effect within each year.  The heat
effect is important as it captures the variability in the amount of time
athletes are on the starting blocks before the gun goes off.  Thus, we can try
to reduce this variability by accounting for it in the model through the heat
effect.
Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. 
%\eds{I was talking to my husband about this. Could the heat effect be justified 
%to capture the variability in the amount of time between the ``set'' and the 
%gun going off (assuming this is different from heat to heat)? E.g., quicker 
%gunshots after the ``set'' could lead to slower reaction times (or the reverse), 
%but this effect would be shared across all athletes in the same heat.}
%\of{Yes I agree with this and updated the language in the text.}
In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function,
$v_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $\delta_{i/j}$ is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

@@

Table~\ref{tab:Clusrankresults} shows the results generated from six 
\eds{three?}
\of{Theres now 6 results in the table, if we remove the asymptotic results then
it should be 3}
tests in R.
We use a \texttt{clusWilcox.test()} with \texttt{exact = TRUE} and 
\texttt{B = 1000} indicating that the random permutation size was chosen to be 
1000, as discussed previously in Section~\ref{sec:rank}.  
%As the sample sizes are small, 65 and 49, we use \texttt{exact = TRUE} as
%that is preferred.  
%In both cases, 
The national versus international comparison
produced lower p values than the 2019 versus 2022 comparison, with the
national vs international permutation test producing a p value of 
$2.2\cdot10^{-16}$, which is the smallest possible value achievable in R.  This
is a highly significant test result and shows that athletes who competed at the
2022 World Track and Field Championships were on average, significantly faster
at reacting than they were a few months prior.  For the 2019 versus 2022
same athlete comparison with permutation argument, the p value was 
$2.9\cdot10^{-3}$, which is significant at $\alpha = 0.01$.  The result of
comparing 2022 and 2023 was very similar to the above result, as the p value
from the permutation test was $3.0\cdot10^{-3}$.  Taken together, these three
tests help show that athletes competing at the 2022 World Track and Field
Championships had faster reaction times than when they had competed previously.

\eds{not sure the \texttt{exact = FALSE} results need to be reported here or in
Table 4 when we say we use \texttt{exact = TRUE} here and in the previous section.}
\of{I am not going to talk about the asymptotic results but will show them in
table.I changed the exact=TRUE and exact = FALSE text to say permutation and 
asymptotic}
these

\eds{need to comment on 2022 vs 2023 results}
\of{Resolved}

@@

A summary of the two effect model is shown in Table~\ref{tab:Gamma_parameters}
for the pooled datasets and shows the standard error of the year
effect increases when including 2022.  The two effect model with venue and heat
effect was the best model as evidenced by a decrease in the AIC compared to the
other models. For the entire data, the AIC for the venue effect only model was
$-2023.5$, the AIC for the heat effect only model $-2070.7$, and the AIC for the 
two effect model, which we will present throughout this section, was $-2075.7$. 
Thus, the two effect model is superior and will be presented throughout the rest
of the paper.
\of{How specific do I need to be in this section? Will people who read this paper
run the code to see the p value or should I say it directly and how do I say it?}
\eds{If you are doing a likelihood ratio test, I suggest including the p-value 
and the degrees of freedom associated with the test.}
\of{Professor Yan recommended not even mentioning the likelihood ratio so
instead I will just talk about the AIC.}

\begin{table}
  \centering
  \caption{Estimated parameters for the two effect GLMM on various data sets. SE
  is short for standard error. \eds{data in table needs to be revisited and table
	needs to be referenced within the text. Also, it needs to be clarified 
	somewhere that ``Old Pooled'' excludes 2022 and ``All Pooled'' includes 2022.}
  \of{Resolved}} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & AIC & $\sigma_h^2$ & $\sigma_v^2$ & $\alpha$ & $\delta_{i/j}$ \\
      \midrule
      Year Effect Only & $-2023.5$ & N/A & 0.048 & $-1.857$ & 0.152\\
      Heat Effect Only & $-2070.7$ & 0.073 & N/A & $-1.890$ & 0.135\\
      Full model       & $-2075.7$ & 0.057 & 0.040 & $-1.880$ & 0.134\\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

@@

This paper does not consider data from the women's 100 meter hurdle, and although
that was initially because of concerns over reaction time differences in men and
women, exploratory analysis showed that 2022 was at best only
slightly faster relative to other championships going back to 2003 
%\eds{check year}. \eds{this seems to contradict the Robert Johnson conclusion in 
%the intro, unless he only considered sprint and not hurdles?} \of{Johnson's
%results considered the mens 100m, mens 110m hurdles, womens 100m but not the 
%womens 100m hurdles.} 
Likewise, the rank-based comparisons resulted in p values
that were not significant. 
\eds{could that be because of even smaller sample sizes?}
\of{Perhaps? I rank the pemutation test for the women's data and the
p value was roughly 0.8.}
  The lack of evidence from the women's data makes it 
harder to conclude that the sensors used at the 2022 championships were faulty, 
although it is unknown if the sensors for the men's 110 meter hurdle and women's
100 meter hurdle were the same.  
\eds{is any calibration of the sensors needed (e.g. for weight?) that might
also play a role for women vs men. (My husband asked me about this as a possible 
reason for the difference in results for men and women, but he did not know
if any calibration is actually done.)}
\of{I honestly am not sure, do you think that this is worth investigating, or
beyond the scope of this paper? I am not sure if the mechanics of Seiko's
technology is public knowledge.}

It is confusing as to why the reaction time barrier for men's and women's data
is the same despite numerous articles such as \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender} stating differently.  World
Athletics is ignoring differences in gender by setting the reaction time barrier
to be equal. If 0.1 seconds is considered a fair time for women, then 0.1 seconds
could not be considered fair for men, as the probability of observing reaction
times lower than 0.1 would be elevated.  \citet{brosnan2017effects} advocated
for gender based reaction time barriers which seems logical if there are biological
differences in reaction times between men and women.

>>>>

\author{Owen Fiore\\
%   \href{mailto:owen.fiore@uconn.edu}
% {\nolinkurl{owen.fiore@uconn.edu}}\\
  Elizabeth D. Schifano\\
  Jun Yan\\[1ex]
  Department of Statistics, University of Connecticut\\
}
\date{}

@@

Devon Allen, an alumnus of the University of Oregon in Eugene, Oregon, was
expected to be a contender for the 2022 World Track and Field Championships 110 
meter hurdle.  After running the event in 12.84 seconds in early June 2022 
(only 0.04 seconds away from the world record \citep{wa2022preview}) and coming 
in third at the United States Track and Field Championships also in Eugene 
in late June, Allen was poised to compete in the 2022 World Track and 
Field Championships in August, which would also be held in Eugene, Oregon.
At the World Championships, Allen made it through the preliminary
heats and semifinals to reach the final heat while posting reaction times of 
0.123 and 0.101 in his preliminary heat and semifinal, respectively.  
His disqualification in the final heat by reacting in 0.099 seconds, just 0.001
seconds faster than the 0.1 threshold set by the International Association of
Athletics Federations (IAAF), was met
with shock from both Allen and fans alike.  NBC, who was broadcasting the
championships in the United States, later uploaded a video to their YouTube
channel that showed the disqualification
(\url{https://www.youtube.com/watch?v=D6NXTMo-1yM}).
The crowd erupted with jeers
when they realized that Allen was disqualified, angry at the result.  Had Allen 
reacted just 0.001 seconds slower he would have been allowed to compete. %and been 
%spared waiting until the next World Track and Field Championships.
%\eds{He is not competing this year, correct? If yes, we may need to rephrase.}
%\of{Correct}
%\eds{removed because he did not compete this year, so did not actually have
% to wait, after all.} 
%\eds{ If it is true he ``retired'' as a runner to play football, we could 
%instead say something like
%\textit{Had Allen reacted just 0.001 seconds slower he would have been allowed 
%to compete before retiring from running to play football.}}
%\of{I don't think he retired, he was competing in races as recently as this past
%June.}
%\eds{Okay, then let's leave it like this.}

@@

The rest of this paper is organized as follows. Section~\ref{sec:Data} details 
how the two types of data were collected, as the signed-rank test as described
in Section~\ref{sec:rank} uses slightly different data than the generalized 
linear mixed-effect model (GLMM) that is further developed in 
Section~\ref{sec:glmm}. Section~\ref{sec:Results} provides the results of the two
analyses and ends by questioning the reaction time barrier.  Lastly, 
Section~\ref{sec:concludingremarks} discusses the impact and limitations of the
paper.

@@

There are two types of data that are explored throughout this paper.  For the
signed-rank tests described in Section~\ref{sec:rank}, the data comes from 
athletes who competed in the 2022 World Championships and another competition 
(2019 World Championships, 2023 World Championships, or 2022 national 
championships). For the GLMM described in Section~\ref{sec:glmm}, data comes
from every world championships from 1999 to 2023. This investigation started 
shortly after the 2022 World Championships, but as the 2023 data has now become 
available, we are pleased to include it in both of our analyses, and it did not
significantly change our results.
\eds{I think you should somehow reference the source of the data here; I saw the
comment below about removing the Appendix and including the sources in the Data 
README file, so assuming the readers will have access to that file, you can 
simply refer to the sources within that file here. I will defer to Prof. Yan's
opinion, however!}

@@

\subsubsection{2022 National Competitions}\label{sec:datanational}
From June 23-26, 2022, the United States held its Track and Field Championships 
at Hayward Field in Eugene, Oregon to decide who to send to the World 
Championships being held at the same venue in August. Thus, we can establish a 
baseline for the four United States athletes who competed in at least one round 
of both events: Trey Cunningham, Daniel Roberts, Grant Holloway, and Devon Allen.   
Every athlete reacted faster in all the World Track and Field Championships 
races compared to the United States Track and Field Championships races. The 
dataset of the four United States athletes is too small to perform any 
meaningful statistical analysis, but the data was expanded to include reaction 
times for 11 more athletes who competed at other national competitions (many of which were 
held from May-July of 2022). This data was difficult to find, as the data is not
centrally located but instead found on different national websites (often 
recorded in the national language without translation). 
%\of{I don't know how to say this}. \eds{is my change correct?} \of{No I was able
%to use a lot even though I couldn't actually read the page. I just tried to find
%110 meter hurdle and then look at a column for reaction times which are always from
%roughly 0.1 to 0.18 seconds.} \eds{is it correct now?} \of{Yes}

@@

@@

While it is theoretically possible that athletes got faster 
between 2019 and 2022 or 2022 and 2023, it is important to 
remember that these are some of the fastest sprinters on the planet, and thus
improving on elite times is not easy.  Once again we repeat a similar procedure
as in Section~\ref{sec:datanational}; we assign each athlete to be a cluster and
within each cluster label times to be from either the 2019, 2022, or 2023 World
Championships. When comparing 2019 and 2022 we look at data from 15
athletes, with two clusters of size two, three clusters of size three, five 
clusters of size four, two clusters of size five, and three clusters of size six.
When comparing 2022 and 2023 we look at data from 22 athletes, with seven 
clusters of size two, three clusters of size three, six clusters of size four,
and six clusters of size five.  
%Although 2022 is presented three times in
%Figure~\ref{fig:RankScatterplots}, the subset of athletes differs each time.
%There may be athletes who competed in 2019 and 2022 and had their national
%championships data available, but they did not compete in 2023, and thus their
%times appear only in the top two graphs. \eds{top and bottom graphs?}
%\eds{also, you will likely need to make Fig 1 smaller as the caption runs off
%the page (even before my edits).}
%\of{It should be smaller now}

@@

\begin{figure}[tbp]
  \centering
  \includegraphics{RankScatterPlots}
  \caption{The top graph shows the reaction times of athletes who competed at 
	least once each at the 2019 and 2022 World Championships. The middle graph 
	shows the reaction times of athletes who competed at least once each at the 
	2022 and 2023 World Championships. The bottom graph shows the reaction times 
	of athletes who competed at least once at their country's
  2022 national championships and the 2022 World Championships.}
	%\eds{check x-axis labels in the bottom plot: last three should be 2022?}
  %\of{I don't think so. Its a comparison of 2022 national vs international so
  %I abbreviated it as Nat. and Inter. is that not ok?}\eds{it's good now}
  \label{fig:RankScatterplots}
\end{figure}

Figure~\ref{fig:RankScatterplots} shows the data that will be analyzed
with the rank-based methods (described in Section~\ref{sec:rank} with results in 
Section~\ref{subsec:Results_Rank}). In each of the three graphs, the reaction 
times for athletes who competed
in the 2022 finals \eds{World Championships?} are compared to their reaction
times at another at which they competed.  
It is worth noting that each instance of the 2022 data is different, because
those who competed at both the 2019 and 2022 World Championships will be
different athletes than those who competed at both the 2022 and 2023 World
Championships. This data was only for athletes who competed
in at least one race at each championship, however it is possible for athletes
to have competed in one race in 2019 and two races in 2022. 
%\eds{did we not already cover the last two sentences in the previous paragraph, 
%or are you trying to say something different here?} \of{yes you are correct. I
%commented out the other one because I think this one flows better.}  
The 2022 World
Championship Heats and Semifinals both had on average lower reaction times than
their respective 2019 races, and the 2022 finals were substantially faster. It
is also worth noting is that the fastest reaction time in the 2022 World
Championships Finals and Semifinals both came from Devon Allen, but the 
difference between 0.101 and 0.099
seconds was the difference in what caused him to be disqualified.

@@

Data was taken from the World Athletics 
and covers the men's 110 meter hurdles from 1999 to 2023.
Although women's data is available, we chose not to include women's data in the
analyses to follow due to the numerous papers citing differences in men and
women's reaction times \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender}.
\eds{you should mention the data source earlier, possible in Section 2 
before Section 2.1 if it applies to Section 2.1.1, and also the sources of the 
national data for 2.1.2 (i.e., pointing to the info in Appendix).}
\of{I want to emphasize the differences in the data so I wanted to try and
compartmentalize everything that was different. We chose not to include women's
data for the GLMM because of cited papers and chose not to include women's
data for the signed-rank because the results were not highly significant. So if
I want to talk about that in section 2, I feel like I would have to explicitly
justify the exclusion of the women's data which then goes back to the
conversation above. Also Prof Yan recommended I remove the appendix, so now the
sources are in the Data README.}
\eds{Got it.  See earlier comment in 2.1; I will defer to Prof. Yan.}
For the creation of the GLMM model we focus on the reaction times during the 
semifinal and final heats only, as reaction times from preliminary heats are 
often not as fast as those in later heats \citep[e.g.,][]{zhang2021correlation}.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics{Boxplot}
  \caption{The reaction times from 1999 to 2023 for the men's 110 meter hurdle.}
  \label{fig:Boxplot}
\end{figure}

@@

With our small sample size, the asymptotic normal distribution may not be
reliable, so we also use 1000 random permutations to simulate the null 
distribution of the test statistic.  
%\eds{simulate the null distribution of the test statistic?}
%\of{Resolved}
This method is available from the \texttt{clusWilcox.test()} function
with \texttt{method = `ds'} (for \underline{D}atta and \underline{S}atten) and 
\texttt{exact = TRUE} from R package
\texttt{clusrank} \citep{jiang2020wilcoxon}.

\subsection{GLMM}\label{sec:glmm}
Based on an exploratory analysis using the GLMM model for the reaction times
from the World Championships, the we chose a gamma mixed-effect model as our 
final model. As suggested by the log-likelihood and the Akaike Information
Criterion (AIC), the normal error model on the log-scale was found to be
much inferior to the gamma model which better captures the skewness of the
data. Two random effects were found useful: one capturing the venue effect in
each year and the other capturing the heat effect within each year.  The heat
effect is important as it captures the variability in the amount of time
athletes are on the starting blocks before the gun goes off.  Thus, we can try
to reduce this variability by accounting for it in the model through the heat
effect.
Let $Y_{ijk}$ be the reaction time of observation~$k$ in heat~$j$ in year~$i$.
Conditional on a venue effect $v_i$ of year~$i$ and a heat effect
$h_{i/j}$ nested within each year~$i$, $Y_{ijk}$ is modeled by a
gamma distribution. 
%\eds{I was talking to my husband about this. Could the heat effect be justified 
%to capture the variability in the amount of time between the ``set'' and the 
%gun going off (assuming this is different from heat to heat)? E.g., quicker 
%gunshots after the ``set'' could lead to slower reaction times (or the reverse), 
%but this effect would be shared across all athletes in the same heat.}
%\of{Yes I agree with this and updated the language in the text.}
In the classic generalized linear model (GLM) sense,
$Y_{ijk}$ has conditional mean $E(Y_{ijk} | v_i, h_{i/j}) > 0$ and a
dispersion parameter~$\phi > 0$. The conditional mean is characterized by
\[
g\{E(Y_{ijk} | v_i, h_{i/j})\} = \alpha + v_i + h_{i/j},
\]
where $g$ is a known link function,
$v_i$ is normally distributed with mean zero and
variance~$\sigma_v^2$, and $\delta_{i/j}$ 
\eds{$h_{i/j}$ or $\delta_{i/j}$? Table 2 uses $\delta_{i/j}$ (are you reporting
the BLUP for this? If yes, why not also report the BLUP of $v_i$ in Table 2?)}
is normally distributed with mean
zero and variance~$\sigma_h^2$.
Although \citet{lo2015idlink} suggested that the gamma 
family with an identity link works well for reaction time in cognitive
psychological research, the logarithm
link function worked better for our data and additionally ensures the positivity
of the conditional mean. The gamma mixed-effect model can be fit with the 
\texttt{glmer()} function in R package \texttt{lme4} \citep{lme4}.

@@

Table~\ref{tab:Clusrankresults} shows the results generated from both the 
permutation (exact) and asymptotic rank-based tests for the three comparisons, 
as discussed previously in Section~\ref{sec:rank}.    
%six \eds{three?}
%\of{Theres now 6 results in the table, if we remove the asymptotic results then
%it should be 3}
%tests in R.
%We use a \texttt{clusWilcox.test()} with \texttt{exact = TRUE} and 
%\texttt{B = 1000} indicating that the random permutation size was chosen to be 
%1000, as discussed previously in Section~\ref{sec:rank}.  
%As the sample sizes are small, 65 and 49, we use \texttt{exact = TRUE} as
%that is preferred.  
%In both cases, 
The national versus international comparison
produced lower p values than the 2019 versus 2022 comparison, with the
national vs international permutation test producing a p value of 
$2.2\cdot10^{-16}$, which is the smallest possible value achievable in R.  
\eds{Prof. Yan, is this possible based on 1000 permutations?} This
is a highly significant test result and shows that athletes who competed at the
2022 World Track and Field Championships were on average, significantly faster
at reacting than they were just a few months prior.  For the 2019 versus 2022
same athlete comparison with permutation test, the p value was 
$2.9\cdot10^{-3}$, which is significant at $\alpha = 0.01$.  The result of
comparing 2022 and 2023 was very similar to the above result, as the p value
from the permutation test was $3.0\cdot10^{-3}$.  Taken together, these three
tests help show that athletes competing at the 2022 World Track and Field
Championships had faster reaction times than when they had competed previously.

%\eds{not sure the \texttt{exact = FALSE} results need to be reported here or in
%Table 4 when we say we use \texttt{exact = TRUE} here and in the previous section.}
%\of{I am not going to talk about the asymptotic results but will show them in
%table.I changed the exact=TRUE and exact = FALSE text to say permutation and 
%asymptotic}
%these
%
%\eds{need to comment on 2022 vs 2023 results}
%\of{Resolved}

@@

A summary of the two effect model is shown in Table~\ref{tab:Gamma_parameters}
for the pooled datasets and shows the standard error of the year
effect increases when including 2022.
\eds{this is not what is currently being shown in
Table~\ref{tab:Gamma_parameters}; also, consider changing ``Full Model'' to 
``Year and Heat Effects''}  
The GLMM with both venue
%two effect model with venue 
\eds{let's try to be consistent with using either
`year' or `venue' in the text, and not both, except when you introduce the intuition behind its use-- it doesn't make a difference to 
me which one you use} 
and heat random
effects was the best model as evidenced by a decrease in the AIC compared to the
other models with only one random effect. %For the entire data \eds{??}, the 
%AIC for the venue effect only model was
%$-2023.5$, the AIC for the heat effect only model $-2070.7$, and the AIC for the 
%two effect model, which we will present throughout this section, was $-2075.7$. 
Thus, results from the GLMM with both venue and heat random effects %is superior and 
will be presented throughout the rest
of the paper.
%\of{How specific do I need to be in this section? Will people who read this paper
%run the code to see the p value or should I say it directly and how do I say it?}
%\eds{If you are doing a likelihood ratio test, I suggest including the p-value 
%and the degrees of freedom associated with the test.}
%\of{Professor Yan recommended not even mentioning the likelihood ratio so
%instead I will just talk about the AIC.}

\begin{table}
  \centering
  \caption{Estimated parameters for the two effect GLMM on various data sets. SE
  is short for standard error. %\eds{data in table needs to be revisited and table
	%needs to be referenced within the text. Also, it needs to be clarified 
	%somewhere that ``Old Pooled'' excludes 2022 and ``All Pooled'' includes 2022.}
  %\of{Resolved}
	\eds{This table is not showing what you have described here or in the main text.}} 
  \begin{tabular}{c c c c c c}
      \toprule
      Dataset & AIC & $\sigma_h^2$ & $\sigma_v^2$ & $\alpha$ & $\delta_{i/j}$ \\
      \midrule
      Year Effect Only & $-2023.5$ & N/A & 0.048 & $-1.857$ & 0.152\\
      Heat Effect Only & $-2070.7$ & 0.073 & N/A & $-1.890$ & 0.135\\
      Full model       & $-2075.7$ & 0.057 & 0.040 & $-1.880$ & 0.134\\
      \bottomrule
  \end{tabular}
  \label{tab:Gamma_parameters}
\end{table}

@@

This paper does not consider data from the women's 100 meter hurdle, and although
that was initially because of concerns over reaction time differences in men and
women, exploratory analysis showed that 2022 was at best only
slightly faster relative to other championships going back to 2003 
Likewise, the rank-based comparisons resulted in p values
that were not significant. 
%\eds{could that be because of even smaller sample sizes?}
%\of{Perhaps? I rank the pemutation test for the women's data and the
%p value was roughly 0.8.}
  The lack of evidence from the women's data makes it 
harder to conclude that the sensors used at the 2022 championships were faulty, 
although it is unknown if the sensors for the men's 110 meter hurdle and women's
100 meter hurdle were the same.  On the other hand, this may be related to 
another debated question regarding whether the reaction time barrier should be 
the same for men and women.
\eds{please check my text addition}  
%\eds{is any calibration of the sensors needed (e.g. for weight?) that might
%also play a role for women vs men. (My husband asked me about this as a possible 
%reason for the difference in results for men and women, but he did not know
%if any calibration is actually done.)}
%\of{I honestly am not sure, do you think that this is worth investigating, or
%beyond the scope of this paper? I am not sure if the mechanics of Seiko's
%technology is public knowledge.}
%\eds{let's not open Pandora's box then}
It is confusing as to why the reaction time barrier for men and women 
is the same despite numerous articles \citep[e.g.,][]{lipps2011implications, 
babicc2009reaction, panoutsakopoulos2020gender} stating differently.  
\eds{The following sentences should be softened a bit, in my opinion, because
it is not clear (at least to me) that these cited authors are any more correct 
than those who advocate equal reaction time barriers for men and women. Please
check my changes.}
According to these authors, World Athletics is ignoring differences in gender by 
setting the reaction time barrier to be equal. 
If there are gender differences and 0.1 seconds is considered a fair time for 
women, then 0.1 seconds could not be considered fair for men, as the probability 
of observing reaction times lower than 0.1 would be elevated.
\citet{brosnan2017effects} advocated for gender based reaction time barriers 
which seems logical if there are biological
differences in reaction times between men and women.
